### 2018-09-28 paper/paper.tex
Update formal analysis results

These qualitative observations can be used to formulate precise statistical hypothesis tests.  Many tests come to mind, but perhaps the simplest and most transparent are based on nothing more than the molecule-by-molecule signs of the ambiguity indexes. Whereas ignoring the actual values of the indexes is inefficient in terms of information, an possibly also in the strict statistical sense, tests based on signs require very few assumptions and are, therefore, more robust to model misspecification. All of the p-values that we will report are based on the hypergeometric distribution, which arises as follows.

We are given a population of $M$ molecules, each with a binary outcome measure $B_m\in\{-1,+1\}$, $m=1,\ldots,M$. There are two subpopulations of interest: the first $M_1$ molecules, making up population 1, and the next $M_2$ molecules, which make up population 2. (So $M_1+M_2=M$.) We observe $n_1$ plus values in population 1 and $n_2$ in population 2 
\begin{align}
n_1 & = \#\big\{m\in\{1,2,\ldots,M_1\}:B_m=+1\big\} \\
n_2 & = \#\big\{m\in\{M_1+1,M_1+2,\ldots,M\}:B_m=+1\big\}
\end{align}
We suspect that population 1 has less than its share of plus ones, meaning that the $n_1+n_2$ population of plus ones was not randomly distributed among the $M$ molecules. To be precise, let $N$ be the number of plus ones that appear from a draw, without replacement, of $M_1$ samples from $B_1,\ldots,B_M$. Under the null hypothesis, $H_o$, $n_1$ is a sample from the hypergeometric distribution on $N$:
\begin{equation}
\label{eqn:null}
\mathbb{P}\{N=n\} = \frac{\binom{M_1}{n}\binom{M_2}{n+1+n_2-n}}
{\binom{M}{n_1+n_2}}\quad \max\{0,n_1+n_2-M_2\}\leq n 
\leq \min\{n_1+n_2,M_1\}
\end{equation}
The alternative hypothesis, $H_a$, is that $n_1$ is too small to be consistent with $H_o$, leading to a left-tail test with p-value $\mathbb{P}\{N\leq n_1\}$ (which can be computed directly or using a statistical package, e.g. {\em hygecdf} in MatLab).

@@

\subsubsection*{Single-entity RNA Molecules versus Protein-RNA Complexes}
Consider an RNA molecule, $m$, selected from one of the seven families in our data set, with primary structure $p$ and secondary structure $s$, computed by comparative analysis.
Given only the T-S ambiguity index of $m$ (i.e. given only $d_\text{T-S}(p,s)$), 
how accurately could we classify $m$ as being unbound (i.e. from the Group I or Group II Introns) as opposed to bound (i.e. from one of the five families tmRNA, SRP RNA, RNase P, 16s rRNA or 23s rRNA)? 
The foregoing exploratory analysis suggests constructing a classifier that declares a molecule to be `unbound' when $d_\text{T-S}(p,s)$ is small, e.g. $d_\text{T-S}(p,s)<t$, where the threshold $t$ governs the familiar trade off between rates of ``true positives'' (an unbound $m$ is declared `unbound') and ``false positives'' (a bound $m$ is declared `unbound').
Large values of $t$ favor low rates of false positives at the price of low rates of true positives, whereas small values of $t$ favor high rates of true-positives at the price of high rates of false positives. Since for each molecule $m$ we have both the correct classification (unbound or bound) and the statistic $d$, we can estimate the ROC performance of our threshold classifier by plotting the empirical values of the pair, (\# false positives,\# true positives), for each value of $t$. The ROC curve for the two-category (unbound versus bound) classifier based on thresholding $d_\text{T-S}(p,s)<t$ is shown in the upper-left panel of Figure \ref{fig:UnboundVSBound}.
Also shown is the estimated area under the curve (AUC=0.81), which has a convenient and intuitive interpretation, as it is equal to the probability that for two randomly selected molecules, $m$ from the unbound population and $m^\prime$ from the bound population, the T-S ambiguity index of $m$ will be smaller than the T-S ambiguity index of $m^\prime$.

@@

As mentioned earlier, we can also associate a traditional p-value to the problem of separating unbound from bound molecules, based on the T-S ambiguity indexes. We consider only the signs (positive or negative) of these indexes, and then test whether there are fewer than expected positive indexes among the unbound, as opposed to the bound, population. This amounts to computing $\mathbb{P}\{N\leq n_1\}$ from the hypergeometric distribution---Eq (\ref{eqn:null}). The relevant statistics can be found in Table \ref{table:NumberPositive}, 
under the column labels {\bf \#mol's} and $\bm \#d_\text{\bf T-S}>0$. Specifically, 
$M_1=116+34=150$ (number of unbound molecules),
$M_2=404+346+407+279+59=1495$ (number of bound molecules), $n_1=50+8=58$ (number of positive T-S indexes among unbound molecules) and $n_2=368+269+379+210+53=1279$.
The resulting p-value, $1.2\cdot 10^{-34}$, is essentially zero, meaning that the positive T-S indexes are not distributed proportional to the sizes of the unbound and bound populations, which is by now obvious, in any case. To repeat our caution, small p-values conflate sample size with effect size, and for that reason we have chosen additional ways, using permutations as well as classifications, to look at the data.

@@

The comparative secondary structure of an RNA molecule, when combined with its primary structure, can be used to construct a measure---the T-S ambiguity index---which distinguishes unbound from bound molecules with good accuracy. Can the same can be said for the D-S index?
Yes, albeit with lower accuracy. To demonstrate, we followed the identical procedure, except that we assigned the index $d_\text{D-S}(p,s)$ rather than $d_\text{T-S}(p,s)$ to each molecule.  The ROC curve (with area 0.62) is shown in the lower-left panel of Figure \ref{fig:UnboundVSBound}. The hypergeometric test, based on the thresholded (signed) values of  $d_\text{D-S}$ (positive counts, for each RNA family, can be found in Table \ref{table:NumberPositive}) has p-value $7.3\cdot 10^{-8}$. Finally, the right-hand panels in Figure \ref{fig:UnboundVSBound} mirror the left-hand panels, except that all secondary structures were computed by (approximately) maximizing free energy rather than comparative analysis. The classification results are substantially less convincing and the p-values substantially higher.

>>>>

These qualitative observations can be used to formulate precise statistical hypothesis tests.  Many tests come to mind, but perhaps the simplest and most transparent are based on nothing more than the molecule-by-molecule signs of the ambiguity indexes. Whereas ignoring the actual values of the indexes is inefficient in terms of information, and probably also in the strict statistical sense, tests based on signs require very few assumptions and are, therefore, more robust to model misspecification. All of the p-values that we will report are based on the hypergeometric distribution, which arises as follows.

We are given a population of $M$ molecules, $m=1,\ldots,M$, each with a binary outcome measure $B_m\in\{-1,+1\}$.
There are two subpopulations of interest: the first $M_1$ molecules make up population 1 and the next $M_2$ molecules make up population 2; $M_1+M_2=M$. We observe $n_1$ plus values in population 1 and $n_2$ in population 2 
\begin{align}
n_1 & = \#\big\{m\in\{1,2,\ldots,M_1\}:B_m=+1\big\} \\
n_2 & = \#\big\{m\in\{M_1+1,M_1+2,\ldots,M\}:B_m=+1\big\}
\end{align}
We suspect that population 1 has less than its share of plus ones, meaning that the $n_1+n_2$ population of plus ones was not randomly distributed among the $M$ molecules. To be precise, let $N$ be the number of plus ones that appear from a draw, without replacement, of $M_1$ samples from $B_1,\ldots,B_M$. Under the null hypothesis, $H_o$, $n_1$ is a sample from the hypergeometric distribution on $N$:
\begin{equation}
\label{eqn:null}
\mathbb{P}\{N=n\} = \frac{\binom{M_1}{n}\binom{M_2}{n+1+n_2-n}}
{\binom{M}{n_1+n_2}}\quad \max\{0,n_1+n_2-M_2\}\leq n 
\leq \min\{n_1+n_2,M_1\}
\end{equation}
The alternative hypothesis, $H_a$, is that $n_1$ is too small to be consistent with $H_o$, leading to a left-tail test with p-value $\mathbb{P}\{N\leq n_1\}$ (which can be computed directly or using a statistical package, e.g. {\em hygecdf} in MatLab).

@@

\subsubsection*{Single-entity RNA Molecules versus Protein-RNA Complexes}
Consider an RNA molecule, $m$, selected from one of the seven families in our data set, with primary structure $p$ and secondary structure $s$, computed by comparative analysis.
Given only the T-S ambiguity index of $m$ (i.e. given only $d_\text{T-S}(p,s)$), 
how accurately could we classify $m$ as being unbound (i.e. from the Group I or Group II Introns) as opposed to bound (i.e. from one of the five families tmRNA, SRP RNA, RNase P, 16s rRNA or 23s rRNA)? 
The foregoing exploratory analysis suggests constructing a classifier that declares a molecule to be `unbound' when $d_\text{T-S}(p,s)$ is small, e.g. $d_\text{T-S}(p,s)<t$, where the threshold $t$ governs the familiar trade off between rates of ``true positives'' (an unbound $m$ is declared `unbound') and ``false positives'' (a bound $m$ is declared `unbound').
Small values of $t$ favor low rates of false positives at the price of low rates of true positives, whereas large values of $t$ favor high rates of true-positives at the price of high rates of false positives. Since for each molecule $m$ we have both the correct classification (unbound or bound) and the statistic $d$, we can estimate the ROC performance of our threshold classifier by plotting the empirical values of the pair
\[
  \text{(\# false positives,\ \# true positives)}
\]
for each value of $t$. The ROC curve for the two-category (unbound versus bound) classifier based on thresholding $d_\text{T-S}(p,s)<t$ is shown in the upper-left panel of Figure \ref{fig:UnboundVSBound}. 
Also shown is the estimated area under the curve (AUC=0.81), which has a convenient and intuitive interpretation, as it is equal to the probability that for two randomly selected molecules, $m$ from the unbound population and $m^\prime$ from the bound population, the T-S ambiguity index of $m$ will be smaller than the T-S ambiguity index of $m^\prime$.

@@

As mentioned earlier, we can also associate a traditional p-value to the problem of separating unbound from bound molecules, based on the T-S ambiguity indexes. We consider only the signs (positive or negative) of these indexes, and then test whether there are fewer than expected positive indexes among the unbound, as opposed to the bound, population. This amounts to computing $\mathbb{P}\{N\leq n_1\}$ from the hypergeometric distribution---Eq (\ref{eqn:null}). The relevant statistics can be found in Table \ref{table:NumberPositive}, 
under the column labels {\bf \#mol's} and $\bm{\#d_\text{\bf T-S}>0}$. Specifically, 
$M_1=116+34=150$ (number of unbound molecules),
$M_2=404+346+407+279+59=1495$ (number of bound molecules), $n_1=50+8=58$ (number of positive T-S indexes among unbound molecules) and $n_2=368+269+379+210+53=1279$.
The resulting p-value, $1.2\cdot 10^{-34}$, is essentially zero, meaning that the positive T-S indexes are not distributed proportional to the sizes of the unbound and bound populations, which is by now obvious, in any case. To repeat our caution, small p-values conflate sample size with effect size, and for that reason we have chosen additional ways, using permutations as well as classifications, to look at the data.

@@

The comparative secondary structure of an RNA molecule, when combined with its primary structure, can be used to construct a measure---the T-S ambiguity index---which distinguishes unbound from bound molecules with good accuracy. Can the same can be said for the D-S index?
Yes, albeit with lower accuracy. To demonstrate, we followed the identical procedure, except that we assigned the index $d_\text{D-S}(p,s)$ rather than $d_\text{T-S}(p,s)$ to each molecule.  The ROC curve (with area 0.65) is shown in the lower-left panel of Figure \ref{fig:UnboundVSBound}. The hypergeometric test, based on the signs of the D-S indexes
(positive counts, for each RNA family, can be found in Table \ref{table:NumberPositive}) has p-value $7.3\cdot 10^{-8}$. Finally, the right-hand panels in Figure \ref{fig:UnboundVSBound} mirror the left-hand panels, except that all secondary structures were computed by (approximately) maximizing free energy rather than comparative analysis. The classification results are substantially less convincing and the p-values substantially higher.
